Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/splitting.tex
Remove references to "correctness" and "syntatically separated".

In this section, we present a {\em splitting algorithm} that
statically converts a term in \lang\ into a pair of monostaged terms
that individually correspond to the first and second stage
computations of the original \lang\ term. Due to important differences
in structure, we describe both the splitting correctness condition and
splitting algorithm for \bbonem\ terms separately from that of
\bbtwo\ terms.  By construction, \bbonep\ terms are monostage and do
not require splitting.

\textbf{Correctness condition for \bbtwo\ term splitting.}  As
suggested informally in \ref{sec:example}, splitting a \lang\ program
(a \bbtwo\ term) should yield two monostage terms where evaluating the
second term (corresponding to second stage computations) on the output
of the first term (first stage computations), yields the same result
as direct evaluation of the original \lang\ program.  More precisely,
for any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e
q$), $e$ splits (via the new relation~$\splittwosym$) into the
syntactically separated terms $\pipeS p l r$ where $\reduce p b$, and
$[b/l]r$ is the same as $q$ up to an equivalence. The term $p$
generated by the split (called the {\em precomputation}, due to its
role producing the inputs for subsequent second stage computation) is
a monostage term representing all the first-stage computations in
$e$. The term $l.r$ (called the {\em resumer}) is a monostage term
representing all the second-stage computations in $e$.  We refer to
the result of the precomputation ($b$) the {\em boundary value} of the
split since it represents the data communicated between the first and
second stage computations.  The process of splitting \bbtwo\ terms
(and the relationship of these operations to the outputs of direct
\lang\ evaluation) is provided at right in \ref{fig:splittingSummary}.
% and described in detail in \ref{sec:split-two}.

@@

\textbf{Correctness condition for \bbonem\ term splitting.} Defining
the correctness requirements for splitting \bbonem\ terms is
challenging because unlike \bbtwo\ terms (which reduce to monostage
residuals that can be directly compared to resumers), \bbonem\ terms
reduce to residual tables and partial values that represent
computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
%For example, the \bbonem\ term:
%\begin{lstlisting}
%(next {1+2}, grnd{3+4})
%\end{lstlisting}
%reduces (via $\redonesym$) to the residual table and partial value:
%\begin{lstlisting}
%[yhat |-> 1+2] (next {yhat}, grnd{7})
%\end{lstlisting}

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired
monostaged outputs of splitting, we choose to define a translation,
called {\em masking}, written $\rtab \xi v \vsplito \mval i q$, which
converts residual tables and partial values into a syntactically
separated {\em masked value}.  The first part of the masked value,
$i$, is a monostage value holding the result of all first stage
components of $v$.  The second component, $q$, is a residual
representing all second stage computations in $\xi$ and $v$. (A precise definition of masking is given in \ref{sec:masking}.)

@@

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, the precomputation $p$ resulting from splitting
\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
the resumer; but the combined term $c$ resulting from splitting
\bbonem\ terms reduces to a tuple containing both the first stage {\em
  immediate result} $i$ and the boundary value $b$ (motivating the
name {\em combined terxm}).

>>>>

In this section, we present a {\em splitting algorithm} that
statically converts a term in \lang\ into a pair of monostaged terms
that individually correspond to the first and second stage
computations of the original \lang\ term. Due to important differences
in structure, we describe both the invariants and algorithm for
splitting \bbonem\ terms separately from that of \bbtwo\ terms.  By
construction, \bbonep\ terms are monostage and do not require
splitting.

\textbf{\bbtwo\ term splitting.}  As suggested informally in
\ref{sec:example}, splitting a \lang\ program (a \bbtwo\ term) should
yield two monostage terms where evaluating the second term
(corresponding to second stage computations) on the output of the
first term (first stage computations), yields the same result as
direct evaluation of the original \lang\ program.  More precisely, for
any $\coltwo e A$ which reduces to a residual $q$ (via $\diatwo e q$),
$e$ splits (via the new relation~$\splittwosym$) into a pair of terms
$\pipeS p l r$ where $\reduce p b$, and $[b/l]r$ is the same as $q$ up
to an equivalence. The term $p$ generated by the split (called the
{\em precomputation}, due to its role producing the inputs for
subsequent second stage computation) is a monostage term representing
all the first-stage computations in $e$. The term $l.r$ (called the
{\em resumer}) is a monostage term representing all the second-stage
computations in $e$.  We refer to the result of the precomputation
($b$) the {\em boundary value} of the split since it represents the
data communicated between the first and second stage computations.
The process of splitting \bbtwo\ terms (and the relationship of these
operations to the outputs of direct \lang\ evaluation) is provided at
right in \ref{fig:splittingSummary}.
% and described in detail in \ref{sec:split-two}.

@@

\textbf{\bbonem\ term splitting.} Defining the requirements for
splitting \bbonem\ terms is challenging because unlike \bbtwo\ terms
(which reduce to monostage residuals that can be directly compared to
resumers), \bbonem\ terms reduce to residual tables and partial values
that represent computations at both stages.
%One complication arises from the fact that \lang\ terms at world \bbonem\ produce multistage output.
%For example, the \bbonem\ term:
%\begin{lstlisting}
%(next {1+2}, grnd{3+4})
%\end{lstlisting}
%reduces (via $\redonesym$) to the residual table and partial value:
%\begin{lstlisting}
%[yhat |-> 1+2] (next {yhat}, grnd{7})
%\end{lstlisting}

To resolve the mismatch between the multi-stage computations
represented by residual tables and partial values and the desired
monostaged outputs of splitting, we choose to define a translation,
called {\em masking}, written $\rtab \xi v \vsplito \mval i q$, which
converts residual tables and partial values into a {\em masked value}.
The first part of the masked value, $i$, is a monostage value holding
the result of all first stage components of $v$.  The second
component, $q$, is a residual representing all second stage
computations in $\xi$ and $v$. (A precise definition of masking is
given in \ref{sec:masking}.)

@@

Note that since $\coltwo e A$ reduces to an entirely second-stage
residual, its first-stage computations only exist to generate input
for second stage computations.  In contrast, since $\colmix e A$
reduces to a multi-stage residual table and partial value, its first
stage computations serve to both generate the value used as input for
the second stage {\em and also} produce a first-stage value.
Correspondingly, the precomputation $p$ resulting from splitting
\bbtwo\ terms reduces only to a boundary value $b$ which is passed to
the resumer; but the combined term $c$ resulting from splitting
\bbonem\ terms reduces to a tuple containing both the first stage {\em
  immediate result} $i$ and the boundary value $b$ (motivating the
name {\em combined term}).
